---
title: "Test_Solution"
author: "Wenjie Lan"
date: "2025-03-11"
output: html_document
---
```{r, warning=FALSE, packages}
library(dirichletprocess)
library(ggplot2)
library(dplyr)
library(tidyr)
set.seed(2)
```

## Easy Test

### Univariate Model for Faithful Dataset

```{r}
data("faithful")
its <- 500

# Standardize 
df_faithful = scale(faithful$waiting)
dp_faithful <- DirichletProcessGaussian(df_faithful)
dp_faithful <- Fit(dp_faithful, its)
plot(dp_faithful)
plot(dp_faithful, data_method="hist")
```

```{r}
data("iris")

# Standardize 
df_iris = scale(iris[,1:4])
dp_iris <- DirichletProcessMvnormal(df_iris, numInitialClusters = nrow(iris))
dp_iris = Fit(dp_iris, 1000)
# Plot
plot(dp_iris)

```
## Medium

- First Step: Generate random data from a lognormal distribution mixture model
- Second Step: Fit and simulate data from DP model
- Last Step : Sample data from posterior distribution,and report 95% CI

**other tasks**

- Explore hyperparameter in Prior: alpha and report its effect on the number of clusters.

- Explore different prior distribution choices effect on the alpha chain

$$
y = 0.2lognormal(0,0.2)+0.6lognormal(2,0.2)+0.2lognormal(3,0.2)
$$
### Sample and simulate data
```{r}
# Sample from mixlognormal
n = 1000
component = sample(0:2, size = n, replace =TRUE, prob = c(0.2,0.6,0.2))
df_lnorm = numeric(n)
df_lnorm[component == 0] = rlnorm(sum(component == 0), meanlog = 0, sdlog = 0.2)
df_lnorm[component == 1] = rlnorm(sum(component == 1), meanlog = 2, sdlog = 0.2)
df_lnorm[component == 2] = rlnorm(sum(component == 2), meanlog = 3, sdlog = 0.2)

# Standardize data and record parameters
mean_original <- attr(scale(df_lnorm), "scaled:center")
sd_original <- attr(scale(df_lnorm), "scaled:scale")
df_lnorm = scale(df_lnorm)


hist(df_lnorm, breaks=30, probability=TRUE,
     main="Simulated Lognormal Mixture Data",
     xlab="Value")
lines(density(df_lnorm), col="red")
```

### Compare number of clusters under different alpha priors

$$
\alpha \sim Gamma(a,b)
$$

with a is the shape parameter and b is the ratio parameter which  our certainty towards alpha. We expect higher alpha generates more clusters.

```{r}
# low alpha prior
dp_low_alpha <- DirichletProcessGaussian(df_lnorm, alphaPrior = c(1, 2))
dp_low_alpha <- Fit(dp_low_alpha, 2000)

# Default mod alpha prior
dp_mod_alpha <- DirichletProcessGaussian(df_lnorm, alphaPrior = c(2, 4))
dp_mod_alpha <- Fit(dp_mod_alpha, 2000)

# high alpha prior
dp_high_alpha <- DirichletProcessGaussian(df_lnorm, alphaPrior = c(5, 2))
dp_high_alpha <- Fit(dp_high_alpha, 2000)
```

```{r}
cat("Number of clusters"," \n")
cat("low alpha:", length(unique(dp_low_alpha$clusterLabels)), "\n")
cat("moderate alpha:", length(unique(dp_mod_alpha$clusterLabels)), "\n")
cat("high alpha:", length(unique(dp_high_alpha$clusterLabels)), "\n")
```


### Calculate 95% CI of posterior Distribution and Parameters

We calculate the 95% CI for posterior distribution under the low alpha prior set with $\alpha = 1,\beta = 2$

```{r}
# hist(df_lnorm, breaks = 30)
plot(dp_low_alpha)
```

$$df_{lnorm} = \frac{df_{unscaled}-Mean}{Sd}$$
```{r}
x <- seq(min(df_lnorm), max(df_lnorm), length.out = 1000)

# Posterior frame
df_pos <- PosteriorFrame(dp_low_alpha, x, 1000)

# Unscale posterior x and density
df_pos$x <- df_pos$x * sd_original + mean_original
density_cols <- c("Mean", "X5.", "X95.")
df_pos[density_cols] <- lapply(df_pos[density_cols], `/`, sd_original)

# True density 
trueFrame <- data.frame(
  x = df_pos$x,
  y = 0.2 * dlnorm(df_pos$x, meanlog = 0, sdlog = 0.2) + 
      0.6 * dlnorm(df_pos$x, meanlog = 2, sdlog = 0.2) + 
      0.2 * dlnorm(df_pos$x, meanlog = 3, sdlog = 0.2)
)

# Sample from Posterior Distribution
samples <- sample(df_pos$x, size = 5000, replace = TRUE, prob = df_pos$Mean)
quantiles <- quantile(samples, probs = c(0.025, 0.5, 0.975))

# Plot posterior and true density
ggplot() +
  geom_histogram(aes(x = samples, y = ..density..), bins = 50, fill = "gray", alpha = 0.5) +
  geom_ribbon(data = df_pos, aes(x = x, ymin = X5., ymax = X95.), fill = "red", alpha = 0.2) + 
  geom_line(data = df_pos, aes(x = x, y = Mean), colour = "red") + 
  geom_line(data = trueFrame, aes(x = x, y = y), colour = "blue") +
  geom_vline(xintercept = quantiles, linetype = "dashed", colour = "black") +
  labs(title = "Posterior Distribution, True Density, and Sampled Histogram",
       x = "Value",
       y = "Density")

cat("Credential Interval of  Posterior Samples:","[",quantiles[0],",",quantiles[2],"]")
```

### Compare the alpha chain under different alpha priors

```{r}
alphaFrame <- data.frame(Chain1 = dp_low_alpha$alphaChain, Chain2 = dp_mod_alpha$alphaChain, Chain3 = dp_high_alpha$alphaChain, Iter=seq_len(2000))
alphaFrame %>% gather(Chain, Value, -Iter) -> alphaFrameTidy

ggplot(alphaFrameTidy, aes(x=Iter, y=Value, colour=Chain))+
  geom_line()+
  scale_color_discrete(labels = c("Low Alpha", "Moderate Alpha", "High Alpha")) +
  labs(title = "Trace Plot of Alpha Chains")
  
```

## Hard

The dirichletprocess package currently implements Dirichlet process mixture models using Gaussian, Beta and Weibull kernels. Now we define a Dirichlet process mixture model (DPMM) in which each mixture component is a multivariate logistic distribution. The logistic distribution has heavier tails, which often increases the robustness of analyses based on it compared with using the normal distribution. By modelling data with a mixture of logistic components, the model can identify the data that do not fit in clusters well.

### Model Set Up: DP with multivariate logistic distribution as its component

Suppose we have pbserved data: $y_1, y_2, ..., y_n$ with $y_i \in R^d$

- Likelihood: $y_i\sim Logistic(\theta_i)$ where $\theta_i = (\mu_i,s_i)$

- Prior:$\theta_i \sim G,$ $G \sim DP(\alpha,G_0)$

$$\mu \sim N_d(m_0, S_0) \quad and \quad s \sim Inverse-Gamma(a_0,b_0)$$ 
Since the logistic distribution is not conjugate to the usual choices of priors, we use MH-MC to update component parameters.

  
```{r}
# Likelihood function
Likelihood.mvlogistic <- function(mdObj, x, theta) {
  nComp <- dim(theta[[1]])[3]  
  nObs <- nrow(x)
  d <- ncol(x)
  likelihoods <- matrix(0, nrow = nObs, ncol = nComp)
  
  for (i in 1:nComp) {
    mu_i <- theta[[1]][, , i] 
    s_i <- theta[[2]][, , i]  
    likelihoods[, i] <- apply(x, 1, function(row) {
      prod(exp(-(row - as.vector(mu_i)) / as.vector(s_i)) /
             (as.vector(s_i) * (1 + exp(-(row - as.vector(mu_i)) / as.vector(s_i)))^2))
    })
  }
  return(likelihoods)
}

# Prior draw function
PriorDraw.mvlogistic <- function(mdObj, n = 1) {
  priorParameters <- mdObj$priorParameters
  d <- length(priorParameters$mu0)
  
  # Draw mu: multinorm
  mu <- array(0, dim = c(d, 1, n))
  for (i in 1:n) {
    mu[, , i] <- matrix(mvtnorm::rmvnorm(1, mean = priorParameters$mu0, sigma = priorParameters$S0), ncol = 1)
  }
  
  # Draw s: inverse gamma
  s <- array(0, dim = c(d, 1, n))
  for (i in 1:n) {
    s_vec <- numeric(d)
    for (j in 1:d) {
      s_vec[j] <- 1 / rgamma(1, shape = priorParameters$a0, rate = priorParameters$b0)
    }
    s[, , i] <- matrix(s_vec, ncol = 1)
  }
  
  theta <- list(mu = mu, s = s)
  return(theta)
}

# Posterior draw 
PosteriorDraw.mvlogistic <- function(mdObj, x, n = 1, n_iter = 1000, burn_in = 500, ...) {
  priorParameters <- mdObj$priorParameters
  d <- ncol(x)
  
  # Initialize parameters
  # set mu to sample mean
  # Set s to sample standard deviation
  current_mu <- colMeans(x)
  current_s <- apply(x, 2, sd)
  current_s[current_s == 0] <- 1
  current_log_s <- log(current_s)
  
  log_posterior <- function(mu, log_s) {
    s <- exp(log_s)
    # Compute log likelihood
    log_lik <- sum(apply(x, 1, function(row) {
      val <- sum( -(row - mu) / s - log(s) - 2 * log(1 + exp(-(row - mu) / s)) )
      if (!is.finite(val)) return(-Inf)
      return(val)
    }))
      
    # Priors, assume iid
    log_prior_mu <- sum(dnorm(mu, mean = priorParameters$mu0, 
                              sd = sqrt(diag(priorParameters$S0)), log = TRUE))
    a0 <- priorParameters$a0; b0 <- priorParameters$b0
    log_prior_s <- sum(a0 * log(b0) - lgamma(a0) - a0 * log_s - b0 / exp(log_s))
    
    total <- log_lik + log_prior_mu + log_prior_s
    if (!is.finite(total)) total <- -Inf
    return(total)}
  
  chain_mu <- matrix(0, nrow = n_iter, ncol = d)
  chain_log_s <- matrix(0, nrow = n_iter, ncol = d)
  chain_mu[1, ] <- current_mu
  chain_log_s[1, ] <- current_log_s
  current_lp <- log_posterior(current_mu, current_log_s)
  
  # Proposal standard deviations 
  prop_sd_mu <- rep(0.1, d)
  prop_sd_log_s <- rep(0.1, d)
  
  for (iter in 2:n_iter) {
    # tuning parameters – adjust
    proposed_mu <- current_mu + rnorm(d, 0, prop_sd_mu)
    proposed_log_s <- current_log_s + rnorm(d, 0, prop_sd_log_s)
    proposed_lp <- log_posterior(proposed_mu, proposed_log_s)
    
    # reject proposals with infinite proposed_lp  
    if (!is.finite(proposed_lp)) {
      accept <- FALSE
    } else {
      accept_prob <- exp(proposed_lp - current_lp)
      if (is.na(accept_prob)) accept_prob <- 0  # safeguard if NA
      accept <- runif(1) < accept_prob
    }
    
    if (accept) {
      current_mu <- proposed_mu
      current_log_s <- proposed_log_s
      current_lp <- proposed_lp
    }
    
    chain_mu[iter, ] <- current_mu
    chain_log_s[iter, ] <- current_log_s
  }
  
  
  # Compute the posterior parameters: average of the post-burn-in
  final_mu <- colMeans(chain_mu[(burn_in + 1):n_iter, , drop = FALSE])
  final_s  <- exp(colMeans(chain_log_s[(burn_in + 1):n_iter, , drop = FALSE]))
  
  # Return the parameters 3D array
  mu_out <- array(matrix(final_mu, ncol = 1), dim = c(d, 1, n))
  s_out  <- array(matrix(final_s,  ncol = 1), dim = c(d, 1, n))
  return(list(mu = mu_out, s = s_out))
}

Predictive.mvlogistic <- function(mdObj, x, n_draws = 1000, ...) {
  theta_draws <- PosteriorDraw.mvlogistic(mdObj, x, n = n_draws, ...)
  lik <- Likelihood.mvlogistic(mdObj, x, theta_draws)
  pred <- rowMeans(lik)
  return(pred)
}

MvlogisticCreate <- function(priorParameters) { 
  mdObj <- MixingDistribution("mvlogistic", priorParameters, "nonconjugate")
  return(mdObj)
}

DirichletProcessMvlogistic <- function(y,
                                       g0Priors,
                                       alphaPriors = c(2, 4),
                                       numInitialClusters = 1) {
  if (!is.matrix(y)) {
    y <- matrix(y, ncol = length(y))
  }
  
  # set defaults base priors
  if (missing(g0Priors)) {
    d <- ncol(y)
    g0Priors <- list(mu0 = rep(0, d),
                     S0 = diag(d),
                     a0 = 2,   
                     b0 = 2)  
  } 
  
  mdobj <- MvlogisticCreate(g0Priors)
  dpobj <- DirichletProcessCreate(y, mdobj, alphaPriors)
  dpobj <- Initialise(dpobj, numInitialClusters = numInitialClusters)
  
  return(dpobj)
}
```

### Model Testing
use iris as example:

```{r}
data("iris")

# Standardize 
df_iris = scale(iris[,1:4])
dp_iris <- DirichletProcessMvlogistic(df_iris, numInitialClusters = nrow(iris))
dp_iris = Fit(dp_iris, 1000)
# Plot
plot(dp_iris)
```

## Reference
- http://127.0.0.1:25612/library/dirichletprocess/doc/dirichletprocess.pdf
- https://cran.r-project.org/web/packages/dirichletprocess/dirichletprocess.pdf